---
title: "Credit card default prediction"
output: html_notebook
---

# Credit Card Default Prediction

Project: Predict whether a customer will default on their credit card
Feb.13th 2024
Selina Wang, Eileen Hu, Alex Huang, Jeffrey Wu, Ziao Xiong

1. Introduction

1.1 Problem description and Objective
(literature review qfehwgrjbkfndlwjaofihgdui)

The goal of this project is to predict whether or not a customer will default on a credit card payment due next month. more context awhgrukfeljdoak

Given a collection of client records in Taiwan, including customer demographics, history of payment, and the case of default payment.

1.2 Data description
The dataset is acquired from Kaggle: https://www.kaggle.com/datasets/uciml/default-of-credit-card-clients-dataset?resource=download

This dataset contains information on default payments, demographic factors, history of repayment status, and bill statements of credit card clients in Taiwan from April 2005 to September 2005. It contains records of 30000 customers with 25 features. Below is a description of them:

- ID: ID of each client
- LIMIT_BAL: Amount of given credit in NT dollars (includes individual and family/supplementary credit)
- SEX: Gender (1=male, 2=female)
- EDUCATION: (0=others, 1=graduate school, 2=university, 3=high school, 4=others, 5=unknown, 6=unknown)
- MARRIAGE: Marital status (0=others, 1=married, 2=single, 3=others)
- AGE: Age in years
Scale of PAY_0, PAY_2, PAY_3, PAY_4, PAY_5, and PAY_6: -2 = no consumption, -1 = paid in full, 0 = use of revolving credit (paid minimum only), 1 = payment delay for one month, 2 = payment delay for two months, …, 8 = payment delay for eight months, 9 = payment delay for nine months and above
- PAY_0: Repayment status in September, 2005 
- PAY_2: Repayment status in August, 2005  
- PAY_3: Repayment status in July, 2005  
- PAY_4: Repayment status in June, 2005  
- PAY_5: Repayment status in May, 2005  
- PAY_6: Repayment status in April, 2005  
All amount of bill statement are shown in New Taiwan dollar (NT dollar)
- BILL_AMT1: Amount of bill statement in September, 2005  
- BILL_AMT2: Amount of bill statement in August, 2005  
- BILL_AMT3: Amount of bill statement in July, 2005  
- BILL_AMT4: Amount of bill statement in June, 2005  
- BILL_AMT5: Amount of bill statement in May, 2005  
- BILL_AMT6: Amount of bill statement in April, 2005  
- PAY_AMT1: Amount of previous payment in September, 2005  
- PAY_AMT2: Amount of previous payment in August, 2005  
- PAY_AMT3: Amount of previous payment in July, 2005  
- PAY_AMT4: Amount of previous payment in June, 2005  
- PAY_AMT5: Amount of previous payment in May, 2005  
- PAY_AMT6: Amount of previous payment in April, 2005  
- default.payment.next.month: Default payment (0 = no, 1 = yes)


2. Exploratory Data Analysis

```{r}
ccard = read.csv("/Users/selina.wang/Desktop/Credit_card_default_prediction/UCI_Credit_Card.csv")
attach(ccard)
str(ccard)
```

Explore the numerical variables: LIMIT_BAL and AGE
```{r}
par(mfrow=c(1,2))

boxplot(LIMIT_BAL~default.payment.next.month, xlab = "Defaulter", col = "steelblue")
boxplot(AGE~default.payment.next.month, xlab = "Defaulter", col = "steelblue")

par(mfrow=c(1,1))
```

Explore the categorical variables: SEX, EDUCATION, MARRIAGE
```{r}
library(ggplot2)
library(dplyr)

categorical_features <- c('SEX', 'EDUCATION', 'MARRIAGE')
ccard_cat <- ccard[categorical_features]
ccard_cat$Defaulter <- ccard$default.payment.next.month

# Replace values with labels
ccard_cat$SEX <- factor(ccard_cat$SEX, labels = c("Male", "Female"))
ccard_cat$EDUCATION <- factor(ccard_cat$EDUCATION, levels = c(1, 2, 3, 4), labels = c("Graduate school", "University", "High school", "Others"))
ccard_cat$MARRIAGE <- factor(ccard_cat$MARRIAGE, levels = c(1, 2, 3), labels = c("Married", "Single", "Others"))

# Plot countplots
for (col in categorical_features) {
  count_plot <- ggplot(ccard_cat, aes(x = !!sym(col), fill = factor(Defaulter))) +
    geom_bar(position = "dodge") +
    labs(title = paste("Countplot of", col),
         x = col,
         y = "Count",
         fill = "Defaulter") +
    theme_minimal()
  
  plot(count_plot)
}
```

Explore the history of repayment status: PAY_0, PAY_2, PAY_3, PAY_4, PAY_5, and PAY_6
```{r}
PAY = c("PAY_0", "PAY_2", "PAY_3", "PAY_4", "PAY_5", "PAY_6")

for (i in PAY) {
  count_plot <- ggplot(ccard, aes(x = !!sym(i), fill = factor(default.payment.next.month))) +
    geom_bar(position = "dodge") +
    labs(title = paste("Countplot of", i),
         x = i,
         y = "Count",
         fill = "Defaulter") +
    theme_minimal()
  
  plot(count_plot)
}
```

Explore the amount of bill statement and previous payment for the past 6 months: 
```{r}
BILL_AMT = c("BILL_AMT1", "BILL_AMT2", "BILL_AMT3", "BILL_AMT4", "BILL_AMT5", "BILL_AMT6")
pairs(ccard[BILL_AMT])

PAY_AMT = c("PAY_AMT1", "PAY_AMT2", "PAY_AMT3", "PAY_AMT4", "PAY_AMT5", "PAY_AMT6")
pairs(ccard[PAY_AMT])
```

Check the correlation between variables
```{r}
library(corrplot)

data <- ccard
correlation_matrix <- cor(data)
corrplot(correlation_matrix)
```
As shown by the plot, the BILL_AMT variables are highly correlated. And, the closer the two months, the more correlated the payment status variables (i.e. PAY_2 and PAY_3). Therefore, in predictive analysis, we are going to use a new predictor to represent the BILL_AMT variables instead of including all of them in the model. Also, we're going to perform the tree models (classification trees, bagging, random forest, boosted trees) to identify the most importance payment status feature before performing logistic regression, LDA, and QDA.

Explore the target variable default.payment.next.month
```{r}
freq <- table(default.payment.next.month)
barplot(freq, col = "steelblue", main = "Target Variable Distribution", ylim=c(0,30000), xlab = "Defaulter", ylab = "Frequency")
```


3. Feature Engineering

Rename the variables
```{r}
column_names <- list(PAY_0 = "PAY_SEPT", PAY_2 = "PAY_AUG", PAY_3 = "PAY_JUL", PAY_4 = "PAY_JUN", PAY_5 = "PAY_MAY", PAY_6 = "PAY_APR", BILL_AMT1 = "BILL_AMT_SEPT", BILL_AMT2 = "BILL_AMT_AUG", BILL_AMT3 = "BILL_AMT_JUL", BILL_AMT4 = "BILL_AMT_JUN", BILL_AMT5 = "BILL_AMT_MAY", BILL_AMT6 = "BILL_AMT_APR", PAY_AMT1 = "PAY_AMT_SEPT", PAY_AMT2 = "PAY_AMT_AUG", PAY_AMT3 = "PAY_AMT_JUL", PAY_AMT4 = "PAY_AMT_JUN", PAY_AMT5 = "PAY_AMT_MAY", PAY_AMT6 = "PAY_AMT_APR", default.payment.next.month = "IsDefaulter"
)

for (old_name in names(column_names)) {
  new_name <- column_names[[old_name]]
  names(ccard)[names(ccard) == old_name] <- new_name
}

names(ccard)
```

Combine and Delete features
```{r}
ccard1 = ccard
attach(ccard1)

# Combine EDUCATION = 0, 4 or 5 to 4
ccard1$EDUCATION = ifelse(ccard1$EDUCATION==0 | ccard1$EDUCATION==4 | ccard1$EDUCATION==5 |ccard1$EDUCATION==6, 4, ccard1$EDUCATION)

# Combine MARRIAGE = 0, 3 to 3
ccard1$MARRIAGE = ifelse(ccard1$MARRIAGE==0 | ccard1$MARRIAGE==3, 3, ccard1$MARRIAGE)

# Combine PAY_month = 4, 5, 6, 7, 8 to 4 for every month
ccard1$PAY_SEPT = ifelse(ccard1$PAY_SEPT==4 | ccard1$PAY_SEPT==5 |ccard1$PAY_SEPT==6 | ccard1$PAY_SEPT==7 | ccard1$PAY_SEPT==8, 4, ccard1$PAY_SEPT)
ccard1$PAY_AUG = ifelse(ccard1$PAY_AUG==4 | ccard1$PAY_AUG==5 |ccard1$PAY_AUG==6 | ccard1$PAY_AUG==7 | ccard1$PAY_AUG==8, 4, ccard1$PAY_AUG)
ccard1$PAY_JUL = ifelse(ccard1$PAY_JUL==4 | ccard1$PAY_JUL==5 |ccard1$PAY_JUL==6 | ccard1$PAY_JUL==7 | ccard1$PAY_JUL==8, 4, ccard1$PAY_JUL)
ccard1$PAY_JUN = ifelse(ccard1$PAY_JUN==4 | ccard1$PAY_JUN==5 |ccard1$PAY_JUN==6 | ccard1$PAY_JUN==7 | ccard1$PAY_JUN==8, 4, ccard1$PAY_JUN)
ccard1$PAY_MAY = ifelse(ccard1$PAY_MAY==4 | ccard1$PAY_MAY==5 |ccard1$PAY_MAY==6 | ccard1$PAY_MAY==7 | ccard1$PAY_MAY==8, 4, ccard1$PAY_MAY)
ccard1$PAY_APR = ifelse(ccard1$PAY_APR==4 | ccard1$PAY_APR==5 |ccard1$PAY_APR==6 | ccard1$PAY_APR==7 | ccard1$PAY_APR==8, 4, ccard1$PAY_APR)

# Delete PAY_AUG, PAY_JUL, PAY_JUN = 1 data # not sure???????????
ccard1 = subset(ccard1, !(PAY_AUG == 1 | PAY_JUL == 1 | PAY_JUN == 1))

# Delete irrelevant variable ID
ccard1 = subset(ccard1, select = -c(ID))
```

Convert categorical variables from int to factor
```{r}
categorical_vars <- c("SEX","EDUCATION","MARRIAGE","PAY_SEPT","PAY_AUG", "PAY_JUL", "PAY_JUN", "PAY_MAY", "PAY_APR", "IsDefaulter")
ccard1[categorical_vars] = lapply(ccard1[categorical_vars], as.factor)
str(ccard1)
```

Create a new variable AVG_BILL_AMT to replace highly correlated BILL_AMT features
```{r}
attach(ccard1)
ccard1$AVG_BILL_AMT = 1/6 * (BILL_AMT_SEPT + BILL_AMT_AUG + BILL_AMT_JUL + BILL_AMT_JUN + BILL_AMT_MAY + BILL_AMT_APR)
ccard1 = subset(ccard1, select = -c(BILL_AMT_SEPT, BILL_AMT_AUG, BILL_AMT_JUL, BILL_AMT_JUN, BILL_AMT_MAY, BILL_AMT_APR))
names(ccard1)
```

Handle missing data
```{r}
dim(ccard1)
ccard1 = na.omit(ccard1)
dim(ccard1)
```

Handle outliers
```{r}
# 3 sigma method

numeric_cols = sapply(ccard1, is.numeric)

valid_rows <- rep(TRUE, nrow(ccard1))

for (i in which(numeric_cols)) {
  mean_col <- mean(ccard1[, i])  
  sd_col <- sd(ccard1[, i])    
  lower_bound <- mean_col - 3 * sd_col
  upper_bound <- mean_col + 3 * sd_col
  # update valid_rows based on 3 sigma bound
  valid_rows <- valid_rows & (ccard1[, i] >= lower_bound & ccard1[, i] <= upper_bound)
}

ccard2 = ccard1[valid_rows,]

dim(ccard2)
```

Perform clustering
```{r}

```

The following models are aimed at predicting the IsDefaulter variable using the rest of the variables in the dataset as predictors.


4. Predictive Analysis

Perform train-test split
```{r}
set.seed(1)

train = sample(nrow(ccard2), nrow(ccard2)*0.7, replace = FALSE)
ccard.train = ccard2[train,]
ccard.test = ccard2[-train,]
```


4.1 Logistic Regression
```{r}
glm.fit = glm(IsDefaulter~., data = ccard.train, family = binomial)

glm.probs = predict(glm.fit, ccard.test, type = "response")

glm.pred = rep(0, nrow(ccard.test))
threshold = 0.5
glm.pred[glm.probs > threshold] = 1

table(glm.pred, ccard.test$IsDefaulter)
logiER = mean(glm.pred != ccard.test$IsDefaulter)
logiER
```


4.2. LDA
```{r}
library(MASS)

lda.fit = lda(IsDefaulter~., data = ccard.train)

lda.pred = predict(lda.fit, ccard.test)
lda.class = lda.pred$class

table(lda.class, ccard.test$IsDefaulter)
ldaER = mean(lda.class != ccard.test$IsDefaulter)
ldaER
```


4.3. QDA
```{r}
qda.fit = qda(IsDefaulter~., data = ccard.train)

qda.pred = predict(qda.fit, ccard.test)
qda.class = qda.pred$class

table(qda.class, ccard.test$IsDefaulter)
qdaER = mean(qda.class != ccard.test$IsDefaulter)
qdaER
```


4.4. Classification Tree
```{r}
library(tree)

tree.ccard = tree(IsDefaulter~., ccard.train)
plot(tree.ccard)
text(tree.ccard, pretty=0)

tree.pred = predict(tree.ccard, ccard.test, type="class")

table(tree.pred, ccard.test$IsDefaulter)
treeER = mean(tree.pred != ccard.test$IsDefaulter)
treeER
```


4.5. Bagging
# probably overfitted, use CV
```{r}
# Tuning the hyperparameter ntree

library(randomForest)
set.seed(1)

testER = c()
ntreeList = seq(250,500,50)
for (ntree in ntreeList) {
  bag.ccard = randomForest(IsDefaulter~., data = ccard.train, mtry = ncol(ccard.train)-1, ntree = ntree, importance=FALSE)
  bag.pred = predict(bag.ccard, ccard.test)
  print(mean(bag.pred != ccard.test$IsDefaulter))
  testER = append(testER, mean(bag.pred != ccard.test$IsDefaulter))
}

plot(ntreeList, testER)
```

```{r}
bestntree = 400
baggingER = testER[4] # hard coding?????????
```


4.6. Random Forest
# probably overfitted, use CV

Tune the hyperparameter mtry
```{r}
# usually mtry = sqrt(numPredictors) gives the best performance

set.seed(1)
testER = c()
mtryList = seq(2,6,1)
for (mtry in mtryList) {
  rf.ccard = randomForest(IsDefaulter~., data = ccard.train, mtry = mtry, ntree = bestntree, importance=FALSE)
  rf.pred = predict(rf.ccard, ccard.test)
  print(mean(rf.pred != ccard.test$IsDefaulter))
  testER = append(testER, mean(rf.pred != ccard.test$IsDefaulter))
}
plot(mtryList, testER)
```

```{r}
bestmtry = 4
rfER = testER[3]
```

Tune the hyperparameter ntree
```{r}
set.seed(1)
testER = c()
ntreeList = seq(250,500,50)
for (ntree in ntreeList) {
  rf.ccard = randomForest(IsDefaulter~., data = ccard.train, mtry = bestmtry, ntree = ntree, importance=FALSE)
  rf.pred = predict(rf.ccard, ccard.test)
  print(mean(rf.pred != ccard.test$IsDefaulter))
  testER = append(testER, mean(rf.pred != ccard.test$IsDefaulter))
}
plot(ntreeList, testER)
```

```{r}
bestntree = 400
rfER = testER[4]
```

Build the best model, and show the importance of predictors
```{r}
set.seed(1)
rf.ccard = randomForest(IsDefaulter~., data = ccard.train, mtry = bestmtry, ntree = bestntree, importance=FALSE)
rf.pred = predict(rf.ccard, ccard.test)
rfER = mean(rf.pred != ccard.test$IsDefaulter)
rfER
  
varImpPlot(rf.ccard)

sorted_indices = order(importance(rf.ccard)[,1], decreasing = TRUE)
importance(rf.ccard)[sorted_indices, ]
```


4.7. Boosted Trees
```{r}
# Tuning shrinkage
library(gbm)

set.seed(1)
threshold = 0.5
testER = c()
shrinkageList = seq(0.05,0.30,0.05)
for (shrinkage in shrinkageList) {
  boost.ccard = gbm(IsDefaulter~., data = ccard.train, distribution = "bernoulli", n.trees = 10, interaction.depth=4, shrinkage = shrinkage)

  boost.probs = predict(boost.ccard, newdata = ccard.test, n.trees = 10)
  boost.pred = ifelse(boost.probs > threshold, 1, 0)
  
  table(boost.pred, ccard.test$IsDefaulter)
  mean(boost.pred == ccard.test$IsDefaulter)
}
plot(ntreeList, testER)



boost.ccard = gbm(IsDefaulter~., data = ccard.train, distribution = "bernoulli", n.trees=1000, interaction.depth=4, shrinkage = 0.1)

boost.probs = predict(boost.ccard, newdata = ccard.test, n.trees = 1000)
threshold = 0.5
boost.pred = ifelse(boost.probs > threshold, 1, 0)

table(boost.pred, ccard.test$IsDefaulter)
mean(boost.pred == ccard.test$IsDefaulter)
```

```{r}
# Tuning interaction.depth


```


```{r}
library(xgboost)

ccard.train <- lapply(ccard.train, function(x) if(is.factor(x)) as.numeric(x) else x)

# Prepare the data
x_train <- ccard.train[, -which(names(ccard.train) == "IsDefaulter")]
x_train <- as.matrix(x_train)
y_train = ccard.train$IsDefaulter
x_test <- ccard.test[, -which(names(ccard.test) == "IsDefaulter")]
x_test <- as.matrix(x_test)
y_test = ccard.test$IsDefaulter

# Train the boosted tree model
dtrain <- xgb.DMatrix(data = as.matrix(x_train), label = y_train)
params <- list(objective = "binary:logistic")
boost_model <- xgboost(params = params, data = dtrain, nrounds = 10)

# Make predictions on the test dataset
dtest = xgb.DMatrix(data = as.matrix(X_test))
predictions <- predict(boost_model, newdata = dtest)

# Convert predicted probabilities to class labels (0 or 1)
predicted_labels <- ifelse(predictions > 0.5, 1, 0)

# Evaluate the predictions (e.g., calculate accuracy)
accuracy <- mean(predicted_labels == ccard.test$IsDefaulter)
print(paste("Accuracy:", accuracy))
```

```{r}
install.packages("catboost")
library(catboost)

# Set parameters
iterations <- 500
learning_rate <- 0.02
depth <- 12
eval_metric <- "AUC"
random_seed <- 42 # Assuming RANDOM_STATE is defined as 42
bagging_temperature <- 0.2
od_type <- "Iter"
metric_period <- 1 # Assuming VERBOSE_EVAL is defined as 1
od_wait <- 100

# Create CatBoost classifier object
clf <- catboost.train(
  data = as.matrix(train_df[, predictors]), 
  label = train_df[, target], 
  verbose = TRUE,
  iterations = iterations,
  learning_rate = learning_rate,
  depth = depth,
  eval_metric = eval_metric,
  bagging_temperature = bagging_temperature,
  od_type = od_type,
  metric_period = metric_period,
  od_wait = od_wait,
  verbose_eval = TRUE
)
```



4.8. Neural Network 

One-hot encoding categorical variables
```{r}
library(fastDummies)

ccard3 = ccard2
categorical_preds <- c("SEX","EDUCATION","MARRIAGE","PAY_SEPT","PAY_AUG", "PAY_JUL", "PAY_JUN", "PAY_MAY", "PAY_APR")
ccard3 = dummy_cols(ccard2, select_columns = categorical_preds)
ccard3 = ccard3[, !colnames(ccard3) %in% categorical_preds]
colnames(ccard3) = make.names(colnames(ccard3)) # make valid names
ccard3[which(colnames(ccard3) == "SEX_1"):ncol(ccard3)] = lapply(ccard3[which(colnames(ccard3) == "SEX_1"):ncol(ccard3)], as.factor) # convert dummies to factor

str(ccard3)
```

Pre-processing the dataset
```{r}
#remove.packages("caret")
#remove.packages("timechange")
#install.packages("timechange")
#install.packages("caret", dependencies = TRUE)
library(caret)

#str(ccard3)

new_ccard <- preProcess(ccard3, method = "range", rangeBounds = c(0,1))
str(new_ccard)

```

Perform train-test split
```{r}
set.seed(1)

train = sample(nrow(ccard3), nrow(ccard3)*0.7, replace = FALSE)
ccard.train = ccard3[train,]
ccard.test = ccard3[-train,]
```


```{r}
library(neuralnet)

nn.ccard = neuralnet(IsDefaulter~., data = ccard.train, linear.output = F, hidden = 5)
# build a neural network model, linear.output = F: classification model
# hidden = c(4,3): 2 hidden layers with 4 nodes and 3 nodes

summary(nn.ccard)
plot(nn.ccard, rep="best")
# display the neural networks

nn.pred = predict(nn.ccard, ccard.test)

mean(nn.pred == ccard.test$IsDefaulter)
```


5. Model Evaluations

```{r}
# need Boosted tree and Neural network
models = c("Logistic Regression", "LDA", "QDA", "Tree", "Bagging", "Random Forest")
errorRates = c(logiER, ldaER, qdaER, treeER, baggingER, rfER)
barplot(errorRates, names.arg = models, col = "steelblue", main = "Error Rates (log) by Model",
        xlab = "Model", ylab = "Error Rate (log)", log = "y", las=3)

errorRates
```

6. Conclusion


